Linear Mixed Models: Random Intercept

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update examples 9, 10, 13

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

>>>>

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

### 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
update examples 9, 10, 13

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = show,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

>>>>

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

### 2024-05-01 examples/2_Examples_ResidualizedChangeScores.qmd
update examples 9, 10, 13

# Mulitple Regression {.unnumbered}

## Overview

<p>

This example assesses whether height, measured at the 1-Year follow-up (T1) in a subsample of ABCD participants, is predicted by weight and height measured at baseline (T0). A visual inspection is further conducted via a scatterplot to graphically represent the relationship between height at Year_1 and weight at baseline, controlling for height at baseline. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

```{r config}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

### Config Options
#::: blue
#**This code configures knitr code chunk options**

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```
:::

:::

## Results

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to conduct a multiple regression analysis by subtracting each participant's height at T1 from their height at T0. Relevant summary statistics are also provided.

**Compute Multiple Regression Model**
```{r build-model-resid-regression-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split data using the correct eventname values
baseline_data <- df %>% 
  filter(eventname == "baseline_year_1_arm_1") %>% 
  select(src_subject_id, Height_baseline = anthroheightcalc, Weight_baseline = anthroweightcalc)

followup_data <- df %>% 
  filter(eventname == "1_year_follow_up_y_arm_1") %>% 
  select(src_subject_id, Height_followup = anthroheightcalc)

# Merge baseline and follow-up data
merged_data <- baseline_data %>%
  left_join(followup_data, by = "src_subject_id")

# Check the first few rows of the merged data
#head(merged_data)

# Fit the regression model excluding NA values
model <- lm(Height_followup ~ Height_baseline + Weight_baseline, data = merged_data, na.action = na.exclude)

@@

:::

### Model Summary {.tabset .tabset-fade .tabset-pills}

::: blue

A plot to show xxxxx.
```{r model-checking}
model_summary <- summary(model)
print(model_summary)

par(mfrow = c(2, 2))  # Set up a 2x2 plotting area
plot(model)  # Generate diagnostic plots

@@

This regression analysis evaluates how baseline height and weight predict children's height at a 1-year follow-up. The output from our model provides several key statistics:

i. an F-statistic of `r model_summary$fstatistic[1]`;
ii. degrees of freedom of `r model_summary$df[2]`;
iii. parameter estimates for baseline height and weight;
iv. standard errors for these estimates; and
v. p-values for the significance of these estimates.

For every one unit increase in baseline height, there's an associated increase of approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units in the follow-up height, and this effect was statistically significant with a p-value of `r model_summary$coefficients["Height_baseline", "Pr(>|t|)"]`. Similarly, for every one unit increase in baseline weight, there's an associated increase of approximately `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units in the follow-up height, which is also statistically significant with a p-value of `r model_summary$coefficients["Weight_baseline", "Pr(>|t|)"]`. Overall, this model explains a substantial portion of the variance in follow-up height, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 4)`. The overall model is highly significant with a p-value less than `r model_summary$coefficients[1,4]`, indicating that the predictors, collectively, have a significant relationship with the dependent variable.

:::

### Model Plots  {.tabset .tabset-fade .tabset-pills}

::: blue
```{r scatterplot}

# Scatterplot with regression line for Weight_baseline
ggplot(merged_data, aes(x = Weight_baseline, y = Height_followup)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relationship between Baseline Weight and Follow-up Height",
       x = "Baseline Weight",
       y = "Follow-up Height") +
  theme_minimal()

@@

The scatterplot visually depicts the relationship between children's weights at baseline (`r names(merged_data)[3]`) and their heights at Year_1 (`r names(merged_data)[4]`). Each point on the plot represents a child, with their baseline weight plotted on the x-axis and their Year_1 height on the y-axis. The clear positive linear trend, as illustrated by the black regression line, indicates that children who had higher weights at baseline generally had higher heights at Year_1. While further statistical analyses can quantify the strength and direction of this relationship, visually, the data suggests a pronounced positive association between baseline weight and follow-up height.

:::
:::

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

The regression analysis was conducted to predict children's height at the 1-Year follow-up using their baseline height (`r names(merged_data)[2]`) and weight (`r names(merged_data)[3]`). Both predictors were statistically significant. Specifically, for every one-unit increase in baseline height, the height at Year_1 increased by approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units, while holding weight constant. Similarly, for every one-unit increase in baseline weight, the height at Year_1 increased by about `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units, while holding baseline height constant. The overall model explained a substantial `r round(model_summary$adj.r.squared * 100, 2)`% of the variance in height at the 1-Year follow-up, as indicated by the adjusted R-squared value.

:::
:::

>>>>

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    group_by(src_subject_id) %>%
    fill(ehi_y_ss_scoreb, .direction = "downup") %>%
    ungroup() %>%
    filter(ehi_y_ss_scoreb %in% c(1, 2)) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
        ehi_y_ss_scoreb = factor(ehi_y_ss_scoreb, levels = c(1, 2), labels = c("Right", "Left")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        handedness = ehi_y_ss_scoreb,
        height = anthroheightcalc
    )

@@

## Overview
A residualized change score is a statistical approach used to account for the baseline (initial) values of a variable when measuring across two time points. First, a linear regression is conducted with the follow-up score predicted by the baseline score. The residual values from this model can be interpreted as change scores that have been adjusted for the baseline values. In a next step, regression analysis can be applied to determine whether a variable of interest predicts the residualized change score.

This example will examine whether handedness predicts changes in height from baseline to the 1 year follow-up, controlling for initial height measurements, in a sample of participants taking part in the ABCD Study. Initially, a linear regression analysis predicts height measurements at the 1 year follow-up from their baseline values to calculate residualized change scores. Subsequently, a second regression assesses whether handedness (left vs. right) predicts these residualized changes in height measurements. The relationship is also visually represented through a violin plot, illustrating differences in the residualized height values between left-handed and right-handed individuals.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lme4) # Linear mixed-effects models
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, handedness, height) %>%
    mutate(event = factor(event)) %>%
    mutate(handedness = factor(handedness)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(handedness ~ "Handedness", height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

@@

## Results
### Compute Multiple Regression Model
The code below is used to compute a residualized change score by first adjusting participants' height at the 1-year follow-up for their baseline height through a regression analysis. The residuals from this model represent the adjusted height changes that are not explained by their initial measurements. Subsequently, these residualized change scores are used in a simple regression analysis to explore whether handedness predicts the changes in height between the two timepoints. This approach helps to isolate any effect of handedness on height growth, independent of the baseline height values.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
    pivot_wider(
        names_from = event,
        values_from = height,
        names_prefix = "Height_"
    )

# Step 1: Fit the linear model to get residuals
model_baseline <- lm(Height_Year_1 ~ Height_Baseline, data = df_wide, na.action = na.exclude)
df_wide <- df_wide %>%
    mutate(residualized_change = residuals(model_baseline))

# Step 2: Model residualized change scores predicted by handedness
model_residualized_change <- lm(residualized_change ~ handedness, data = df_wide, na.action = na.exclude)

# View the summary of the model to see the effect of handedness
summary(model_residualized_change)

@@

The results from the residualized change score analysis suggest that while individual differences in growth occurs, handedness does not appear to be a determining factor. The model intercept, close to zero (-0.005), is consistent with the expectation that if there is no effect for handedness, there would be no change when adjusting for baseline heights. Further, examination of the model residuals (range = -53.90 to 65.53, se = 2.32) and the negligible r-squared value indicates that handedness does not explain a substantive portion of the variance in residualized height changes.

### Model Plots
```{r}

par(mfrow = c(2, 2)) # Set up a 2x2 plotting area
plot(model_baseline) # Generate diagnostic plots

# Visualize the residualized change scores across different levels of handedness

# Create a violin plot to show the distribution of residualized change scores by handedness
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(RColorBrewer)

# Plotting using ggplot2
ggplot(df_long, aes(x = handedness, y = height, fill = handedness)) +
    geom_violin() +
    geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Residualized Change Score by Handedness",
        x = "Handedness",
        y = "Height Change Score"
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    )

The violin plot visualizes the distribution of residualized change scores in children's heights by their handedness (Right, Left). Each violin shape provides insight into the density of the data at different (residualized) height values, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the overall distributions of the residualized change scores for the height outcome across the two handedness categories are markedly similar.

## Wrapping Up
This analysis evaluated residualized change scores of participants' height over a one-year period, controlling for initial height measurements at baseline. It then utilized a simple regression analysis to investigate handedness as a predictor of these adjusted changes in height. Findings demonstrated a non-significant effect of handedness on the residualized height values, with left-handed individuals showing an average height change increase of .06 inches (r = .77, se = .08, p = .43) more than right-handed individuals. A visual inspection using a violin plot further supported these results, showing overlapping distributions of residualized height changes across handedness categories, indicating no substantial effect.

This approach of using residualized change scores can be particularly valuable when initial measurement levels may influence the outcome. In some context, this analysis can provide a more accurate reflection of change by accounting for baseline values, making it suitable for analyses where measurement of change is a primary aim and where baseline characteristics might otherwise obscure true changes.

### 2024-05-01 examples/_queue/1b_minimalist_Examples_DifferenceScores_SimpleRegression.qmd
update examples 9, 10, 13

---
title: "Difference Scores: Simple Regression"
author: "Biostatistics Working Group"
---

@@

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

>>>>

---
title: "Difference Scores: Simple Regression" ()
author: "Biostatistics Working Group"
---

@@

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

### 2024-05-01 examples/_queue/3a_minimalist_Examples_LinearMixedModels.qmd
update examples 9, 10, 13

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"
---

@@

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(cbcl_scr_syn_external_r) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = cbcl_scr_syn_external_r
    )

df_long <- df_long %>%
  group_by(id) %>%
  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
  ungroup()

@@

## Overview

The Linear mixed model (LMM) with a random intercept and slope extends the traditional fixed-effect linear regression by incorporating random effects for the intercept and slope parameters. This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.

This example will examine trajectories of height obtained across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time. The LMM with a random intercept and slope will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept and slope values.

@@

### Descriptives
```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(cbcl_scr_syn_external_r) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = cbcl_scr_syn_external_r
    )

df_long <- df_long %>%
  group_by(id) %>%
  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
  ungroup()

@@

## Results 
### Compute LMM with Random Intercepts and Slopes
This code fits a random-intercept and random-slope LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept and slope values across participants. Model output is provided below, as well as a brief summary of results.

This model accounts for individual variability in height by including a random intercept and sllope for each subject (id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

@@

model <- lmer(height ~ event + (1 + event | id), data = df_long, REML = T)

# Model with random intercepts only
model_ri_only <- lmer(height ~ event + (1 | id), data = df_long, REML = T)
model_ri_only

# Model with random slopes only, without intercepts
model_rs_only <- lmer(height ~ event + (0 + event | id), data = df_long, REML = T)
model_rs_only

@@

These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The random intercept and slope effects suggests some between-subject variability in particpant's height values at baseline and in rates of change over time. The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.

### Model Plots
The following set of plots are used to facilitate model diagnostics. The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects. The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity. The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.

# 1. Extract the random effects
random_effects <- ranef(model)[[1]]

# 2. Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# 3. hist_plot
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, color = "black", fill = "lightblue") +
    labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
    theme_minimal()

print(hist_plot)

@@

# Generate the plot
ggplot(df_long, aes(x = event, y = height, group = id)) +
    geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
    geom_line(data = df_long, aes(x = event, y = predicted_height, group = 1, color = "Mean Trajectory"), linewidth = 1.5) +
    scale_color_manual(values = c("Individual Trajectories" = "grey", "Mean Trajectory" = "blue")) +
    labs(
        title = "Individual and Group-Mean Height Trajectories",
        x = "Event Name",
        y = "Height (cm)",
        color = "Trajectory Type"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

@@

This plot illustrates the change in participant's height across 3 annual measurement occassions. The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point. These are overlaid with grey lines, which adjust each individual's trajectory unique growth pattern indicated by the fixed effects from the model. The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.

The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights and rates of change in height over tim. On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2. The random effects for the estimated intercept and slope suggest differences in starting heights and rates of change over time, reflecting the variability in growth patterns among the participants.

## Wrapping Up

The random-intercept and random-slope LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate. An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with evidence of individual variability around this estimate.

The LMM framework used in this example offered the ability to account for individual differences in initial height and changes in growth over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects. More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals. This enables a more precise understanding of both the group-level trends and individual-specific trajectories in longitudinal data.

>>>>

---
title: "Linear Mixed Models: Random Intercept"
author: "Biostatistics Working Group"
---

@@

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

@@

## Overview
The Linear mixed model (LMM) with a random intercept is similar to traditional (fixed-effect) linear regression, extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value.

This example will examine trajectories of height obtained across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time. The random-intercept LMM will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept value.

@@

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

@@

## Results
### Compute LMM Model with Random Intercepts
This code fits a radom-intercept LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept values across participants. Model output is provided below, as well as a brief summary of results.

This model accounts for individual variability in height by including a random intercept for each subject (id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

@@

model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = T)

@@

These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The random intercepts effect suggests some between-subject variability in particpant's height values at baseline, with a standard deviation of about 3.03 inches. The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.

### Model Plots
The following set of plots xxxxx.

@@

# Generate the plot
ggplot(df_long, aes(x = event, y = height)) +
    geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
    geom_line(aes(y = predicted_height + individual_intercept, group = id, color = "Adjusted Individual Trajectories"), alpha = 0.7) +
    geom_line(aes(x = event, y = predicted_height, group = 1, color = "Sample Mean Trajectory"), size = 1.5) +
    scale_color_manual(values = c("Individual Trajectories" = "red", "Adjusted Individual Trajectories" = "grey", "Sample Mean Trajectory" = "blue")) +
    labs(
        title = "Individual Trajectories with Fixed Slope and Random Intercepts",
        x = "Event Name",
        y = "Height (inches)",
        color = "Trajectory Type"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

@@

This plot illustrates the change in participant's height across 3 annual measurement occassions. The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point. These are overlaid with grey lines, which adjust each individual's trajectory to start from their unique intercept while following a common growth pattern indicated by the fixed effects from the model. The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.

The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights. On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2. The random effects for individual intercepts suggest differences in starting heights, with a standard deviation of 3.03 inches, reflecting the variability in growth patterns among the participants.

## Wrapping Up
The random-intercept LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate. An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with the rate of change held constant across all participants.

The LMM framework used in this example offered the ability to account for individual differences in initial height and the consistent growth patterns over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects. More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals. This enables a more precise understanding of both the group-level trends and individual-specific trajectories in longitudinal data.
